Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-11-20 analysis/paper/paper.tex
submitted to PLOS

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{fancyhdr, graphicx}
\usepackage[hidelinks]{hyperref}

\fancyhead{\vspace{70pt}}
\renewcommand{\headrulewidth}{1.8pt}

\makeatletter
\def\@maketitle{%
  \newpage
  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
  \null
 \begin{center}
{\LARGE
\textbf\newline{\@title}
}
%\newline

\end{center}
  \par}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\maketitle

@@

\thispagestyle{fancy}

@@

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
Sørensen}\textsuperscript{2}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{3}, \normalsize{Amanda G.
Henry}\textsuperscript{1}\vspace{12pt}

\rule{\textwidth}{1.8pt}\vspace{8pt}

@@

\textsuperscript{\footnotesize{1}}\textit{\small{Department of
Archaeological Sciences, Leiden
University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
of Forensic Medicine, Aarhus
University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
Genootschap Beemster}}

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* \small{b.p.bartholdy@arch.leidenuniv.nl}

@@

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 56\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, and especially
children and women.
\end{abstract}

@@

\vfill

\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-sample-demography-1.pdf}

}

\caption{\label{fig-sample-demography}Overview of sample demography.
Left plot is the first batch and right plot is the replication batch
with 29 of the individuals from the first batch. eya = early young adult
(18-24 years); lya = late young adult (25-34 years); ma = middle adult
(35-49 years); old = old adult (50+ years). Male? = probable male;
Female? = probable female.}

\end{figure}

@@

All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
were calculated with the \textbf{psych} package (Revelle, 2022).

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
(Figure~\ref{fig-auth-plot-batch2}).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-auth-plot-batch2-1.pdf}

}

\caption{\label{fig-auth-plot-batch2}(A) Number of samples in which each
compound was detected in the first and second batch. (B) Quantity (ng)
of each compound extracted from each sample in batch 2. The plot
displays the extracted quantity across the three washes and final
calculus extraction (calc). Each coloured line represents a different
calculus sample. CBD = cannabidiol; CBN = cannabinol; THC =
tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA =
tetrahydrocannabivarin acid.}

\end{figure}

To see if preservation of the skeletal remains had any effect on the
detection of compounds, we compare extracted quantities of compounds to
the various levels of skeletal preservation. Our results from batch 2
suggest that detection of a compound may be linked to the preservation
of the skeleton, with better preservation leading to increased
extraction quantity (Figure~\ref{fig-detection-preservation}A). We also
find a weak positive correlation between the weight of the calculus
sample and the quantity of compound extracted from the calculus
(Figure~\ref{fig-detection-preservation}B).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-detection-preservation-1.pdf}

}

\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2. r = Pearson correlation
coefficient.}

\end{figure}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.85). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
caries and periodontitis (0.52), periodontitis and VOP (0.49),
periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
CMS (0.43). Remaining correlations were weak or absent
(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-polycorr-1.pdf}

}

\caption{\label{fig-polycorr}Plot of the polychoric correlations
(\emph{rho}). Larger circles and increased opacity indicates a stronger
correlation coefficient. OA = osteoarthritis; VOP = vertebral
osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO
= cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic
acid.}

\end{figure}

@@

Nicotine and its principal/main metabolite, cotinine, were strongly
positively correlated, both in concentration and presence/absence in
individuals (Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). The
detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature (Aten et al.,
2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
(Lemmers et al., 2013). There is also documented medicinal use of
nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
headaches, respiratory problems, colds, and drowsiness from around 1780
to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
and cotinine (replicated) in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the
dental calculus through pipe-smoking, as the individual had no visible
pipe notches; more likely the tobacco entered through an alternate mode
of consumption, secondhand smoke, or the aforementioned tobacco-smoke
enema.

Theophylline and caffeine were positively correlated in our samples,
though to a lesser extent than nicotine and cotinine, so we are unable
to determine if they originated from the same source
(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
theophylline have very similar chemical structures, so we expect they
would experience similar rates of incorporation and degradation,
allowing us to interpret the ratio and correlations between the
compounds. Caffeine is present in coffee, tea, and cocoa beans, with
concentrations slightly higher in coffee (Bispo et al., 2002; Chin et
al., 2008; Srdjenovic et al., 2008; Stavric et al., 1988). Theophylline
is present in both coffee beans and tea leaves, but in negligible
quantities (Stavric et al., 1988). It is also a primary metabolite of
caffeine produced by the liver. Given the low correlation, there are
likely multiple sources of caffeine and theophylline in the population,
with tea and coffee being the most obvious.\\
Tea consumption had become widespread in the Netherlands by 1820,
reaching all parts of society (Nierstrasz, 2015, p. 91). Historically,
we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
special occasion occurring from 15.00-20.00 hours, where tea was served
along with the evening bread (Schuijtemaker, 2011). Many households also
owned at least one coffee pot and tea pot (Bouman, 2017). Distinguishing
between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is
present in higher quantities in chocolate compared to caffeine and
theophylline (Alañón et al., 2016; Bispo et al., 2002; Stavric et al.,
1988). However, In addition to oral factors affecting alkaloid uptake in
dental calculus, there is some indication that theobromine does not
preserve well in the archaeological record (Velsko et al., 2017), and
frequent consumption of all three items would be difficult to parse.
Additionally, we do not understand well enough the effect of the burial
on these specific compounds, and the original concentration of the
compounds in plants can be quite variable (King et al., 2017).

Salicylic acid was found in all but one individual in our sample. It can
be extracted from the bark of willow trees, \emph{Salix alba}, and has
long been used for its pain-relieving properties (Bruinsma, 1872, p.
119). It is also present in many plant-based foods (Duthie \& Wood,
2011; Malakar et al., 2017), including potatoes, which were a staple of
the Beemsterpolder diet (Aten et al., 2012). The extracted quantity from
our samples decreased over the three washes, followed by a sharp
increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life
(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
especially with salicylic acid, there is a possibility for the compound
to enter the calculus through contact with the surrounding soil.
Salicylic acid is a very mobile organic acid (Badri \& Vivanco, 2009;
Chen et al., 2001) and the ubiqutous presence in our samples may be
explained by the compound leching into the dental calculus from the
burial environment. We can therefore not confidently rule out
environmental contamination without analysing samples from the
surrounding soil.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,bottom=30mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{lineno}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig}
\else
  \newcommand\figurename{Fig}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%\maketitle

@@

%\thispagestyle{fancy}

\LARGE{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}

@@

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1,2*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{3}, \normalsize{Lambert K.
Sørensen}\textsuperscript{3}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{4}, \normalsize{Amanda G.
Henry}\textsuperscript{1}\vspace{12pt}

%\rule{\textwidth}{1.8pt}\vspace{8pt}

@@

\textsuperscript{\footnotesize{1}}\textit{\small{Department of
Archaeological Sciences, Leiden
University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Faculty
of Mechanical, Maritime and Materials Engineering, Delft University of
Technology}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Department
of Forensic Medicine, Aarhus
University}}\\ \textsuperscript{\footnotesize{4}}\textit{\small{Historisch
Genootschap Beemster}}

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* \small{b.p.bartholdy@tudelft.nl}

@@

\linenumbers

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological lesions and
presence of alkaloids. Dental calculus was analysed using
ultra-high-performance liquid chromatography-tandem mass spectrometry
(UHPLC-ESI-MS/MS). We were able to detect nicotine, cotinine, caffeine,
theophylline, and salicylic acid, suggesting the consumption of tea and
coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches
in the dentition. Nicotine and/or cotinine was present in 56\% of
individuals with at least one visible pipe notch. There is some
influence of skeletal preservation on the detection of alkaloids, with
higher quantities of compounds extracted from well-preserved
individuals, and we observe a positive relationship between weight of
the calculus sample and quantity of detected compounds, as well as
between chronic maxillary sinusitis and the presence of multiple
alkaloids. There are many limitations that will need to be addressed
going forward with this type of analysis; we stress the need for more
systematic research on the consumption of alkaloid-containing items and
their subsequent concentration and preservation in dental calculus, in
addition to how mode of consumption may affect concentrations in the
dentition. Despite the limitations, this preliminary study illustrates
many benefits of using calculus to target a variety of compounds that
could have been consumed as medicine or diet. This method allows us to
directly address specific individuals, which can be especially useful in
individuals that are not always well-documented in historic
documentation, such as rural populations, and especially children and
women.
\end{abstract}

@@

%\vfill

%\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (\textbf{?@fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

\emph{Fig 1 (fig-sample-demography): Overview of sample demography. Left
plot is the first batch and right plot is the replication batch with 29
of the individuals from the first batch. eya = early young adult (18-24
years); lya = late young adult (25-34 years); ma = middle adult (35-49
years); old = old adult (50+ years). Male? = probable male; Female? =
probable female.}

@@

All statistical analysis was conducted in R version 4.3.2 (2023-10-31),
Eye Holes, (R Core Team, 2020). Data wrangling was conducted with the
\textbf{tidyverse} (Wickham et al., 2019) and visualisations were
created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
were calculated with the \textbf{psych} package (Revelle, 2022).

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
(\textbf{?@fig-auth-plot-batch2}).

\emph{Fig 2 (fig-auth-plot-batch2): (A) Number of samples in which each
compound was detected in the first and second batch. (B) Quantity (ng)
of each compound extracted from each sample in batch 2. The plot
displays the extracted quantity across the three washes and final
calculus extraction (calc). Each coloured line represents a different
calculus sample. CBD = cannabidiol; CBN = cannabinol; THC =
tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA =
tetrahydrocannabivarin acid.}

To see if preservation of the skeletal remains had any effect on the
detection of compounds, we compare extracted quantities of compounds to
the various levels of skeletal preservation. Our results from batch 2
suggest that detection of a compound may be linked to the preservation
of the skeleton, with better preservation leading to increased
extraction quantity (\textbf{?@fig-detection-preservation}A). We also
find a weak positive correlation between the weight of the calculus
sample and the quantity of compound extracted from the calculus
(\textbf{?@fig-detection-preservation}B).

\emph{Fig 3 (fig-detection-preservation): (A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2. r = Pearson correlation
coefficient.}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.85). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
caries and periodontitis (0.52), periodontitis and VOP (0.49),
periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
CMS (0.43). Remaining correlations were weak or absent
(\textbf{?@fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

\emph{Fig 4 (fig-polycorr): Plot of the polychoric correlations
(}rho\emph{). Larger circles and increased opacity indicates a stronger
correlation coefficient. OA = osteoarthritis; VOP = vertebral
osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO
= cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic
acid.}

@@

Nicotine and its principal/main metabolite, cotinine, were strongly
positively correlated, both in concentration and presence/absence in
individuals (Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). The
detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature (Aten et al.,
2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
(Lemmers et al., 2013). There is also documented medicinal use of
nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
headaches, respiratory problems, colds, and drowsiness from around 1780
to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
and cotinine (replicated) in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the
dental calculus through pipe-smoking, as the individual had no visible
pipe notches; more likely the tobacco entered through an alternate mode
of consumption, secondhand smoke, or the aforementioned tobacco-smoke
enema.

Theophylline and caffeine were positively correlated in our samples,
though to a lesser extent than nicotine and cotinine, so we are unable
to determine if they originated from the same source
(Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). Caffeine and
theophylline have very similar chemical structures, so we expect they
would experience similar rates of incorporation and degradation,
allowing us to interpret the ratio and correlations between the
compounds. Caffeine is present in coffee, tea, and cocoa beans, with
concentrations slightly higher in coffee (Bispo et al., 2002; Chin et
al., 2008; Srdjenovic et al., 2008; Stavric et al., 1988). Theophylline
is present in both coffee beans and tea leaves, but in negligible
quantities (Stavric et al., 1988). It is also a primary metabolite of
caffeine produced by the liver. Given the low correlation, there are
likely multiple sources of caffeine and theophylline in the population,
with tea and coffee being the most obvious.\\
Tea consumption had become widespread in the Netherlands by 1820,
reaching all parts of society (Nierstrasz, 2015, p. 91). Historically,
we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
special occasion occurring from 15.00-20.00 hours, where tea was served
along with the evening bread (Schuijtemaker, 2011). Many households also
owned at least one coffee pot and tea pot (Bouman, 2017). Distinguishing
between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is
present in higher quantities in chocolate compared to caffeine and
theophylline (Alañón et al., 2016; Bispo et al., 2002; Stavric et al.,
1988). However, In addition to oral factors affecting alkaloid uptake in
dental calculus, there is some indication that theobromine does not
preserve well in the archaeological record (Velsko et al., 2017), and
frequent consumption of all three items would be difficult to parse.
Additionally, we do not understand well enough the effect of the burial
on these specific compounds, and the original concentration of the
compounds in plants can be quite variable (King et al., 2017).

Salicylic acid was found in all but one individual in our sample. It can
be extracted from the bark of willow trees, \emph{Salix alba}, and has
long been used for its pain-relieving properties (Bruinsma, 1872, p.
119). It is also present in many plant-based foods (Duthie \& Wood,
2011; Malakar et al., 2017), including potatoes, which were a staple of
the Beemsterpolder diet (Aten et al., 2012). The extracted quantity from
our samples decreased over the three washes, followed by a sharp
increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life
(\textbf{?@fig-auth-plot-batch2}). It is important to note that,
especially with salicylic acid, there is a possibility for the compound
to enter the calculus through contact with the surrounding soil.
Salicylic acid is a very mobile organic acid (Badri \& Vivanco, 2009;
Chen et al., 2001) and the ubiqutous presence in our samples may be
explained by the compound leching into the dental calculus from the
burial environment. We can therefore not confidently rule out
environmental contamination without analysing samples from the
surrounding soil.
